|

### 2023-06-08 manuscript/paper.Rmd
More text and references added.

Ecological release theory suggests that a prey population’s niche should shift and expand when predation is relaxed or removed (reviewed in @Herrmann2021). 
Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011). Theory and empirical data suggest that a population presented with ecological opportunity following ecological release might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (@Schoener1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022). Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution, 16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>). The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; @Roestietal2023). The lineage adaptively evolved armor loss but also reduction in the means of several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al. in prep). In particular, tooth wear data suggest that the population also began eating more planktonic prey, indicative of increased open water habitat use (Purnell et al. 2007). If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect to observe sexual dimorphism evolution for traits related to habitat use in <i>G. doryssus</i>.

We imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.  Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense to test whether increased sexual dimorphism is associated with ecological release from predators.

@@

A major challenge in this work is inferring the sex of fossil specimens. This typically cannot be done directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020). Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020). We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G. doryssus is part of the extant threespine stickleback species complex (Gasterosteus aculeatus).  Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.

We then combined these sets of data and treat the sex of the fossil specimens as missing and  used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.  Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations performed.

After imputing the sex of the fossil data, each completed data set is then used to fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes

>>>>

Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed, as prey take advantage of new enemy free space (reviewed in @Herrmann2021). Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011). Theory and empirical data suggest that a population presented with ecological opportunity following ecological release might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (@Schoener1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022). Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for phenotypic sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).

We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, finely-resolved, ~16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>). The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020). However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reichen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023). The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al. in prep). In particular, tooth wear data suggest that the population shifted to eat more planktonic prey through time, indicative of increased open water habitat use (Purnell et al. 2007). If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect sexual dimorphism to evolve for traits related to habitat use in <i>G. doryssus</i>.

To test this prediction, we imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.  Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for 16 traits related to swimming, feeding and defense.

@@

A major challenge in this work is inferring the sex of fossil stickleback specimens. Indeed, for most species, sex cannot be detected directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020). Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020). We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G. doryssus is part of the extant threespine stickleback species complex, Gasterosteus aculeatus. We measured the same trait set from a phenotypically diverse set of extant G. aculeatus populations, for whom individual sex was known. We then combined fossil and extant data sets, treated fossil sex as missing data, and used the MICE (@MICE) multiple imputation (@little2002statistical) algorithm, and implemented in R (@R2022language), to impute individual sex of the fossil data set 100 times. We then fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to the data to test for the evolution of sexual dimorphism in each trait.
